\vspace{-2.5em} Spatio-angular transfer functions for fluorescence microscopes\vspace{-1em}

### 2018-03-12 notes/2018-02-23-spatio-angular-kernel/report/report.tex
Update notes for group meeting. Filling gaps on path to OTF.

\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\ro}[1]{\mathbf{\mathfrak{r}}_o}
\providecommand{\so}[1]{\mathbf{\hat{s}}_o}
\providecommand{\rb}[1]{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}[1]{\mathbf{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\bs}[1]{\boldsymbol{#1}} 
\providecommand{\intinf}{\int_{-\infty}^{\infty}}
\providecommand{\fig}[4]{
  % filename, width, caption, label
\begin{figure}[h]
 \captionsetup{width=1.0\linewidth}
 \centering
 \includegraphics[width = #2\textwidth]{#1}
 \caption{#3}
 \label{fig:#4}
\end{figure}
}

@@

We use plain roman type for scalars, e.g., $x, y, z$; bold lowercase roman type
for two-dimensional vectors, e.g., $\mb{r}$; hats for unit vectors, e.g.,
$\mb{\hat{s}}$; bold lowercase gothic type for three-dimensional vectors, e.g.,
$\mb{\mathfrak{r}}$; and bold capital roman type for matrices, e.g.,
$\mb{R}$. We use the real spherical harmonic functions
\begin{align}
  y_l^m(\theta, \phi) =
  \begin{cases}
    \sqrt{2}K_l^m\cos(m\phi)P_l^m(\cos\theta), & m > 0\\
    K_l^0P_l^0(\cos\theta), & m = 0\\
    \sqrt{2}K_l^m\sin(-m\phi)P_l^{-m}(\cos\theta), & m < 0\\
  \end{cases}
\end{align}
where
\begin{align}
  K_l^m = \sqrt{\frac{(2l+1)}{4\pi}\frac{(l-|m|)!}{(l+|m|)!}},
\end{align}
and $P_l^m(x)$ are the associated Legendre polynomials. The $l=0$ and $l=1$
spherical harmonics are given by
\begin{equation}
\begin{aligned}
  y_0^0(\theta, \phi) &= \sqrt{\frac{1}{4\pi}},\\
  y_1^{-1}(\theta, \phi) = \sqrt{\frac{3}{4\pi}}\sin\phi\sin\theta, \hspace{2em} y_1^0(\theta, \phi) &= \sqrt{\frac{3}{4\pi}}\cos\theta, \hspace{2em} y_1^1(\theta, \phi) = \sqrt{\frac{3}{4\pi}}\cos\phi\sin\theta.
\end{aligned}
\end{equation}

\section{General forward model}
Consider a three-dimensional field of oriented fluorescent dipoles. We can
represent the entire field using a function $f(\ro{}, \so{})$ that returns the
number of dipoles at position $\ro{}$ per unit volume oriented in direction
$\so{}$ per unit solid angle. We call $f(\ro{}, \so{})$ the fluorescence
spatio-angular density because it has units of fluorophores per volume per solid
angle.

If each fluorophore in the field fluoresces independently then the measured
intensity is a weighted sum of the intensity contributions from each
fluorophore, and we can model the imaging system with
\begin{align}
  g_i(\rd{}) = \int_{\mathbb{R}^3}d\ro{} \int_{\mathbb{S}^2}d\so{}\ h_i(\rd{}; \ro{}, \so{})f(\ro{}, \so{})\label{eq:forward}
\end{align}
where $g_i(\mb{r_d})$ are the intensities measured by the detector under the
$i$th measurement configuration ($i$ can index polarizer settings or views),
$\mb{r_d}$ is the two-dimensional detector coordinate, and
$h_i(\rd{}; \ro{}, \so{})$ is the spatio-angular point spread function (PSF) of
the $i$th measurement configuration. Our goal is to reconstruct the object
$f(\ro{}, \so{})$ from intensity measurements $g_i(\rd{})$.

We can also use the spatio-angular density to model translating and rotating
fluorophores. As long as each fluorescence emission is an independent event,
then equation \ref{eq:forward} is a valid way to model the imaging system.

\section{Spatio-angular point spread function}
\begin{figure}[h]
 \captionsetup{width=1.0\linewidth}
 \centering
   \centering
   \includegraphics[width = 0.4\textwidth]{../figures/schematic.pdf}
   \caption{Simplified schematic of a single-view fluorescence microscope. The
     object is placed near the focal point of an aplanatic objective lens with
     focal length $f_{o}$ in a medium with refractive index $n_o$. The object is
     parameterized by the 3D position vector $\ro{}$ ($o$ for object) and an
     orientation unit vector $\so{}$. The light emitted by the fluorescent
     object is collected and collimated by the objective lens so that the
     electric fields are purely transverse in the back focal plane. Points in
     the back focal plane are parameterized by a 2D position vector $\rb{}$ ($b$
     for back focal plane). Finally, the tube lens with focal length $f_t$
     refocuses the light onto a detector. Points on the detector are
     parameterized by a 2D position vector $\rd{}$ ($d$ for detector). The back
     focal plane and detector are in a medium with refractive index $n_t$. Note
     that this schematic is not to scale---we consider the case where
     $f_o \ll f_t$.}
   \label{fig:frames_a}
\end{figure}

Figure \ref{fig:frames_a} shows a schematic of the fluorescence microscope that
we are considering with a summary of our notation. We start by following Backer
and Moerner \cite{backer2014} to find the electric field in the back focal plane
due to a single dipole emitter at position $\ro{}$ oriented along $\so{}$ as 
\begin{align}
  \mb{e_b}(\rb{};\ro{}, \so{}) \propto \me^{in_ok(z_o\rho_b)}\sqrt{\frac{1}{\rho_b}}
  \begin{bmatrix}
    \sin^2\phi_b + \rho_b\cos^2\phi_b&\sin\phi_b\cos\phi_b(\rho_b - 1)&-r_b\cos\phi_b\\
    \sin\phi_b\cos\phi_b(\rho_b - 1)&\cos^2\phi_b + \rho_b\sin^2\phi_b&-r_b\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfp}
\end{align}
where $r_b$ is the distance from the center of the back focal plane in units of
$f_o$, we define $\rho_b \equiv \sqrt{1 - r_b^2}$, and $\Pi(x)$ is a boxcar
function that returns 1 when $|x| < 1$ and 0 otherwise. We can understand this
expression term by term---the exponential term accounts for the defocus phase,
the square root term conserves power before and after the objective lens, the
matrix models the dipole emission process and electric field rotation caused by
the objective lens, the vector is the dipole orientation unit vector, and
$\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)$ accounts for the numerical
aperture of the lens with $r_b^{\text{max}} = \frac{f_0}{n_0}\text{NA}$. Notice
that we have omitted the dependence on the transverse position of the
fluorophore $x_o$ and $y_o$. This is because we assume that the objective is
aplanatic, so it displays transverse shift-invariance and the constant phase
factor does not affect our discussion.

We can find the intensity in the detector plane by propagating the electric
field to the detector plane using a scaled Fourier transform then taking the
modulus squared
\begin{align}
  h_i(\rd{}; \ro{}, \so{}) = \left|\int_{\mathbb{R}^2} d\rb{} \mb{e}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t) \rb{} \cdot \rd{}}\right|^2 \label{eq:kernel}.
\end{align}
Following Novotny \cite{nov2006}, we write the Fourier transform in polar
coordinates, evaluate the azimuthal integrals, and write the result concisely in
terms of three radial integrals
\begin{align}
    h_i(\rd{}; \ro{}, \so{}) &= \left|\mb{e}_d(\rd{}; \ro{}, \so{})\right|^2 \label{eq:kernel2}.\\
  \mb{e}_d(\rd{}; \ro{}, \so{}) &\propto
  \begin{bmatrix}
    I_0 + I_2\cos(2\phi_d) & I_2\sin(2\phi_d) & -2iI_1\cos\phi_d\\
    I_2\sin(2\phi_d) & I_0 - I_2\cos(2\phi_d) & -2iI_1\sin\phi_d\\
    0&0&0\\
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}\label{eq:elec}
\end{align}
where $I_0, I_1$ and $I_2$ are
\begin{align}
  I_0(r_d; z_o) &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1+\cos\theta)J_0(k_b r_d\sin\theta f_o/f_t)\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  I_1(r_d; z_o) &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin^2\theta J_1(k_b r_d\sin\theta f_o/f_t)\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  I_2(r_d; z_o) &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1-\cos\theta)J_2(k_b r_d\sin\theta f_o/f_t)\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\}.
\end{align}
By plugging Eq. \ref{eq:elec} into equation Eq. \ref{eq:kernel2} and expanding in
terms of the spherical harmonic functions, we can rewrite the spatio-angular
PSF as
\begin{equation}
  \begin{split}
  h(\rd{}; \ro{}, \so{}) \propto &\left(I_0^2 + 2I_1^2 + I_2^2\right)y_0^0(\so{}) -\frac{2\sqrt{15}}{5}I_0I_2\sin(2\phi_d)y_2^{-2}(\so{})\\ &+ \frac{1}{\sqrt{5}}\left(-I_0^2 + 4I_1^2 - I_2^2\right)y_2^{0}(\so{}) +\frac{2\sqrt{15}}{5}I_0I_2\cos(2\phi_d)y_2^{2}(\so{}). \label{eq:kerna}
\end{split}
\end{equation}

We have chosen to write the spatio-angular PSF in terms of spherical harmonic
functions for two reasons. First, it allows us to express the spatio-angular PSF
very concisely. Instead of considering the point spread function for every
possible dipole orientation, we only need to consider four spatio-angular
PSFs---one for each spherical harmonic. Second, the spherical harmonic functions
form an orthonormal basis for functions on the sphere---a convenient fact that
we will use later.

@@

For now, let's look closer at the spatio-angular PSF in a special case. By
constraining the object to the focal plane ($z_o$ = 0) and applying the paraxial
approximation ($\sin\theta\approx\theta$ and $\cos\theta\approx 1$), the
integrals $I_{0-2}$ can be evaluated in terms of Bessel functions. We can
evaluate $I_0$ and $I_1$ with the help of $\int_0^zdx\ xJ_0(ax) = zJ_1(az)/a$
and $\int_0^zdx\ x^2J_1(ax) = z^2J_2(az)/a$, respectively, and $I_2 = 0$ because
$J_2(x)$ is zero to first order. In this case, the spatio-angular PSF simplifies
to
\begin{align}
      \lim_{\theta_{\text{max}} \ll \pi/2} h(\rd{}; \mathbf{r}_o, z_o=0, \so{}) \equiv h^{(p)}(\rd{}; \mathbf{r}_o, \so{}) \propto \left({I_0^{(p)}}^2 + 2{I_1^{(p)}}^2\right)y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left(-{I_0^{(p)}}^2 + 4{I_1^{(p)}}^2\right)y_2^{0}(\so{})\label{eq:para}
\end{align}
where the integrals evaluate to
\begin{align}
  {I_0^{(p)}} = \left[\frac{2J_1(\tilde{r}_d)}{\tilde{r}_d}\right],
  \hspace{2em}
    {I_1^{(p)}} = \frac{\text{NA}}{n_o}\left[\frac{2J_2(\tilde{r}_d)}{\tilde{r}_d}\right],
  \intertext{and we have substituted}
  \tilde{r}_d = \frac{2\pi\text{NA}\thinspace r_d}{M\lambda},\hspace{2em}
  \text{NA} = n_o\sin\theta_{\text{max}},\hspace{2em}
  M = \frac{n_o}{n_b}\frac{f_t}{f_o}.
\end{align}
We will continue to use an superscipt $(p)$ to mark the terms where we have used
the paraxial approximation and constrained the object to the focal plane.

First, consider the coefficient on the $y_0^0$ spherical harmonic in
Eq. \ref{eq:para}. This coefficient is the point spread function for a angularly
uniform distribution of fluorophores. The first term of the coefficient $I_0^{(p)}$
is the familiar Airy disk that arises from the contribution of dipoles oriented
in the transverse plane, while the second term $I_1^{(p)}$ is a smaller factor that
arises from dipoles oriented outside of the transverse plane. This leads to an
interesting conclusion---a uniform distribution of dipoles has a point spread
function that is slightly wider than an Airy disk even in the paraxial
approximation. The Airy disk is usually derived using paraxial scalar optics
while here we have used paraxial electromagnetic optics. Therefore, we can
consider the second term to be an electromagnetic correction to the Airy
disk. We will quantify this difference in the next section.

@@

\section{Spatio-angular transfer functions}
We define the spatio-angular optical transfer function (OTF) as
\begin{align}
  H_l^m(\bs{\nu}_o) = \int_{\mathbb{R}^3}d\ro{} \int_{\mathbb{S}^2}d\so{}\ h_i(\rd{}; \ro{}, \so{}) y_l^m(\so{})\thinspace\me^{i 2\pi \ro{}\cdot\bs{\nu}_o}. \label{eq:ft}
\end{align}
The spatio-angular OTF measures the ability of the microscope to pass
spatio-angular harmonics---instead of the usual spatial harmonics
$\me^{i 2\pi \ro{}\cdot\bs{\nu}_o}$ we now need consider the spatio-angular
harmonics $y_l^m(\so{})\thinspace\me^{i 2\pi
  \ro{}\cdot\bs{\nu}_o}$. Eq.~\ref{eq:ft} can be interpreted as the
spatio-angular Fourier transform of the spatio-angular PSF.

>>>>

\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mf}[1]{\mathbf{#1}}
\providecommand{\ro}[1]{\mathbf{\mathbf{r}}_o}
\providecommand{\so}[1]{\mathbf{\hat{s}}_o}
\providecommand{\rb}[1]{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}[1]{\mathbf{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\bs}[1]{\boldsymbol{#1}} 
\providecommand{\intinf}{\int_{-\infty}^{\infty}}
\providecommand{\fig}[4]{
  % filename, width, caption, label
\begin{figure}[h]
 \captionsetup{width=1.0\linewidth}
 \centering
 \includegraphics[width = #2\textwidth]{#1}
 \caption{#3}
 \label{fig:#4}
\end{figure}
}

@@

We use plain roman type for scalars, e.g., $x, y, z$; bold lowercase roman type
for two-dimensional vectors, e.g., $\mb{r}$; hats for unit vectors, e.g.,
$\mb{\hat{s}}$; and bold capital roman type for matrices, e.g.,
$\mb{R}$. We use the real spherical harmonic functions
\begin{align}
  y_l^m(\vartheta, \varphi) =
  \begin{cases}
    \sqrt{2}K_l^m\cos(m\varphi)P_l^m(\cos\vartheta), & m > 0\\
    K_l^0P_l^0(\cos\vartheta), & m = 0\\
    \sqrt{2}K_l^m\sin(-m\varphi)P_l^{-m}(\cos\vartheta), & m < 0\\
  \end{cases}
\end{align}
where
\begin{align}
  K_l^m = \sqrt{\frac{(2l+1)}{4\pi}\frac{(l-|m|)!}{(l+|m|)!}},
\end{align}
and $P_l^m(x)$ are the associated Legendre polynomials. The $l=0$ and $l=1$
spherical harmonics are given by
\begin{equation}
\begin{aligned}
  y_0^0(\vartheta, \varphi) &= \sqrt{\frac{1}{4\pi}},\\
  y_1^{-1}(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\sin\varphi\sin\vartheta, \hspace{2em} y_1^0(\vartheta, \varphi) &= \sqrt{\frac{3}{4\pi}}\cos\vartheta, \hspace{2em} y_1^1(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\cos\varphi\sin\vartheta. \label{eq:harmonics}
\end{aligned}
\end{equation}

\section{Spatio-angular point spread functions}
\begin{figure}[h]
 \captionsetup{width=1.0\linewidth}
 \centering
   \centering
   \includegraphics[width = 0.4\textwidth]{../figures/schematic.pdf}
   \caption{Simplified schematic of a single-view fluorescence microscope. The
     object is placed near the focal point of an aplanatic objective lens with
     focal length $f_{o}$ in a medium with refractive index $n_o$. The object is
     parameterized by the 2D position vector $\ro{}$ ($o$ for object) and an
     orientation unit vector $\so{}$. The light emitted by the fluorescent
     object is collected and collimated by the objective lens so that the
     electric fields are purely transverse in the back focal plane. Points in
     the back focal plane are parameterized by a 2D position vector $\rb{}$ ($b$
     for back focal plane). Finally, the tube lens with focal length $f_t$
     refocuses the light onto a detector. Points on the detector are
     parameterized by a 2D position vector $\rd{}$ ($d$ for detector). The back
     focal plane and detector are in a medium with refractive index $n_t$. Note
     that this schematic is not to scale---we consider the case where
     $f_o \ll f_t$.}
   \label{fig:frames_a}
\end{figure}

Figure \ref{fig:frames_a} shows a schematic of the fluorescence microscope that
we are considering with a summary of our notation. We start by following Backer
and Moerner \cite{backer2014} to find the electric field at position $\rb{}$ in
the back focal plane due to a single dipole emitter at position $\ro{}$ oriented
along $\so{}$ as
\begin{align}
  \mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \propto \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}\sqrt{\frac{1}{\rho_b}}
  \begin{bmatrix}
    \sin^2\phi_b + \rho_b\cos^2\phi_b&\sin\phi_b\cos\phi_b(\rho_b - 1)&-\frac{r_b}{f_o}\cos\phi_b\\
    \sin\phi_b\cos\phi_b(\rho_b - 1)&\cos^2\phi_b + \rho_b\sin^2\phi_b&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfp}
\end{align}
where we define $\rho_b \equiv \sqrt{1 - \left(\frac{r_b}{f_o}\right)^2}$, and
$\Pi(x)$ is a boxcar function that returns 1 when $|x| < 1$ and 0 otherwise. We
can understand this expression term by term---the exponential term accounts for
the phase as dipole emitter is moved in object space, the square root term
conserves power before and after the objective lens, the matrix models the
dipole emission process and electric field rotation caused by the objective
lens, the vector is the dipole orientation unit vector, and
$\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)$ accounts for the numerical
aperture of the lens with $r_b^{\text{max}} = \frac{f_0}{n_0}\text{NA}$. We use
a tilde to mark single dipole response functions---we will consider the response
due to a field of dipoles in the next section.

If the tube lens is weakly focusing ($f_o \ll f_t$) then we can find the
electric field in the detector plane by taking the Fourier transform of the
electric field in the back focal plane
\begin{align}
  \mb{\tilde{e}}_d(\rd{}; \ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\mb{e}}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t) \rb{} \cdot \rd{}}. \label{eq:ed}
\end{align}
By isolating the phase term in Eq. \ref{eq:bfp} with
$\mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) = \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
\tilde{\underline{\mb{e}}}_b(\rb{};\ro{}, \so{})$, and plugging into Eq. \ref{eq:ed} we find
that 
\begin{align}
  \mb{\tilde{e}}_d(\rd{} - M\ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\underline{\mb{e}}}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t)\, \rb{} \cdot [\,\rd{} - M\ro{}]}.
\end{align}
where $M = -\frac{n_o}{n_t}\frac{f_t}{f_o}$ is the transverse magnification.  By
writing the electric field in the detector plane in terms of $\rd{} - M\ro{}$,
we have established that the electric field in the detector plane is
\textit{transverse shift-invariant}---a transverse shift of the object creates a
magnified transverse shift of the image. We define coordinates on the detector
centered on the image of the object
$\rd{}' \equiv \rd{} - M\ro{} = r_d'\cos\phi_d'\mh{x} + r_d'\sin\phi_d\mh{y}$,
then we follow Novotny \cite{nov2006} by writing the integrals in polar
coordinates, evaluating the azimuthal integrals, and writing the result
concisely in terms of three radial integrals
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    A + C\cos(2\phi_d') & C\sin(2\phi_d') & -2iB\cos\phi_d'\\
    C\sin(2\phi_d') & A - C\cos(2\phi_d') & -2iB\sin\phi_d'\\
    0&0&0\\
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}\label{eq:elec}
\end{align}
where $A, B$ and $C$ are
\begin{align}
  A(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1+\cos\theta)J_0(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  B(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin^2\theta J_1(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  C(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1-\cos\theta)J_2(k_b r_d'\sin\theta f_o/f_t).%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\}.
\end{align}
We can identify Eq. \ref{eq:elec} as the vector-valued \textit{coherent spread
  function} (CSF) of the microscope. The scalar-valued CSF---sometimes called
the amplitude transfer function---is only applicable to cases where
electromagnetic optics plays no role in the microscope---not true
for oriented samples.

To build our intuition about the CSF, we will rewrite the matrix multiplication
in Eq. \ref{eq:elec} in terms of the spherical harmonics. Notice that the $l=1$
spherical harmonics are the Cartesian components of the unit dipole axis (see
Eq. \ref{eq:harmonics}), so the CSF becomes
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    [A + C\cos(2\phi_d')]y_1^{1}(\so{}) + C\sin(2\phi_d')y_1^{-1}(\so{}) - 2iB\cos\phi_d'y_1^{0}(\so{})\\
    C\sin(2\phi_d')y_1^{1}(\so{}) + [A - C\cos(2\phi_d')]y_1^{-1}(\so{}) - 2iB\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}.
\end{align}
By applying the paraxial approximation ($\sin\theta\approx\theta$ and
$\cos\theta\approx 1$), the integrals $A, B$ and $C$ can be evaluated in terms
of Bessel functions. We can evaluate $A$ and $B$ with the help of
$\int_0^zdx\ xJ_0(ax) = zJ_1(az)/a$ and $\int_0^zdx\ x^2J_1(ax) = z^2J_2(az)/a$,
respectively, and $C = 0$ because $J_2(x)$ is zero to first order. In this
case, the CSF simplifies to
\begin{align}
  \tilde{\mb{e}}^{(p)}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    A^{(p)}y_1^{1}(\so{}) - 2iB^{(p)}\cos\phi_d'y_1^{0}(\so{})\\
    A^{(p)}y_1^{-1}(\so{}) - 2iB^{(p)}\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}\label{eq:paracsf}
\end{align}
where the integrals evaluate to
\begin{align}
  {A^{(p)}} = \frac{2J_1(\tilde{r}_d)}{\tilde{r}_d},
  \hspace{2em}
    {B^{(p)}} = \frac{\text{NA}}{n_o}\left[\frac{2J_2(\tilde{r}_d)}{\tilde{r}_d}\right],
  \intertext{and we have substituted}
  \tilde{r}_d = \frac{2\pi\text{NA}\thinspace r_d}{M\lambda},\hspace{2em}
  \text{NA} = n_o\sin\theta_{\text{max}}.
\end{align}
We will continue to use an superscipt $(p)$ to mark the terms where we have used
the paraxial approximation.

Under the paraxial approximation the electric fields on the detector created by
a single dipole are composed of two parts, a parallel part and a radial
part. The coefficients of the $y_1^0$ spherical harmonic in Eq. \ref{eq:paracsf}
create the radial part of the field---the $z$ component of the dipole generates
a radial field on the detector with respect to the image point. The coefficients
of the $y_1^1$ and $y_1^{-1}$ spherical harmonics create a parallel field in the
detector plane parallel to the $x$ and $y$ components of the
dipole. As the NA increases and the paraxial approximation no longer applies, we
begin to see coupling between the $x(y)$ component of the dipole with the $y(x)$
transverse field.

We can find the intensity in the detector plane due to a single dipole
by taking the modulus squared of the CSF
\begin{align}
  \tilde{h}(\rd{}', \so{}) = \left|\tilde{\mb{e}}_d(\rd{}', \so{})\right|^2 \label{eq:kernel}.
\end{align}
For convenience we keep the CSF written in terms of the spherical harmonics and
use a table of spherical harmonic products (see Appendix \ref{realspherical}) to
calculate the intensity in the detector plane as
\begin{equation}
  \begin{split}
  \tilde{h}(\rd{}', \so{}) \propto &\left(A^2 + 2B^2 + C^2\right)y_0^0(\so{}) -\frac{2\sqrt{15}}{5}AC\sin(2\phi_d')y_2^{-2}(\so{})\\ &+ \frac{1}{\sqrt{5}}\left(-A^2 + 4B^2 - C^2\right)y_2^{0}(\so{}) +\frac{2\sqrt{15}}{5}AC\cos(2\phi_d')y_2^{2}(\so{}). \label{eq:kerna}
\end{split}
\end{equation}
We can identify Eq. \ref{eq:kerna} as the spatio-angular \textit{point spread
  function} (PSF) of the microscope. Writing the spatio-angular PSF in terms of
spherical harmonic functions has two advantages. First, it allows us to express
the spatio-angular PSF very concisely. Instead of considering the point spread
function for every possible dipole orientation, we only need to consider four
spatio-angular PSFs---one for each spherical harmonic. Second, the spherical
harmonic functions form an orthonormal basis for functions on the sphere---a
convenient fact that we will use later.

@@

The spatio-angular PSF under the paraxial approximation is given by
\begin{align}
      \tilde{h}^{(p)}(\rd{}', \so{}) \propto \left({A^{(p)}}^2 + 2{B^{(p)}}^2\right)y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left(-{A^{(p)}}^2 + 4{B^{(p)}}^2\right)y_2^{0}(\so{})\label{eq:para}.
\end{align}
First, consider the coefficient on the $y_0^0$ spherical harmonic in
Eq. \ref{eq:para}. This coefficient is the point spread function for a angularly
uniform distribution of fluorophores. The first term of the coefficient ${A^{(p)}}^2$
is the familiar Airy disk that arises from the contribution of dipoles oriented
in the transverse plane, while the second term ${B^{(p)}}^2$ is a smaller factor that
arises from dipoles oriented outside of the transverse plane. This leads to an
interesting conclusion---a uniform distribution of dipoles has a point spread
function that is slightly wider than an Airy disk even in the paraxial
approximation. The Airy disk is usually derived using paraxial scalar optics
while here we have used paraxial electromagnetic optics. Therefore, we can
consider the second term to be an electromagnetic correction to the Airy
disk. We will quantify this difference in the next section.

@@

\section{Spatio-angular transfer functions}
Consider a thin object that consists of fluorescent dipoles in arbitrary
positions and orientations. We can represent the entire object using a function
$\bs{\mu}(\ro{}, \so{})$ that returns a complex-valued vector for each position
$\ro{}$ and direction $\so{}$. The magnitude of the complex-valued vector is the
magnitude of the dipole moment and the elements of the vector give the orientation
and phase of the radiating dipole moment. Because $\bs{\mu}(\ro{}, \so{})$
includes the relative phases of different points and orientations in the object,
$\bs{\mu}(\ro{}, \so{})$ can represent coupled dipoles that are emitting
partially or completely coherently.

We can find the electric field on the detector created by this object by
multiplying $\bs{\mu}(\ro{}, \so{})$ with the CSF then integrating over all
positions and orientations in the object
\begin{align}
  \mb{e}_d(\rd{}) = \int_{\mathbb{S}^2}d\so{}\int_{\mathbb{R}^2}d\ro{} \ \tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{})\bs{\mu}(\ro{}, \so{}).\label{eq:eforward}
\end{align}
Note that these integrals represent a vector sum---the coherence of the electric
fields radiated by the object can cause cancellations of the fields created on
the detector.

We can simplify this expression by expanding the CSF and the object in terms
of spatio-angular harmonics
\begin{align}
  \tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{}) &= \sum_{l=0}^{\infty} \sum_{m=-l}^l \int_{\mathbb{R}^2}d\bs{\nu}_o{}\, \mb{E}_l^m(\bs{\nu}_o)y_l^m(\so{})\thinspace\me^{i 2\pi (\rd{} - M\ro{})\cdot\bs{\nu}_o},\label{eq:csfexp}\\
    \bs{\mu}(\mb{r}_o, \so{}) &= \sum_{l=0}^{\infty} \sum_{m=-l}^l \int_{\mathbb{R}^2}d\bs{\nu}_o{}\, \mb{M}_l^m(\bs{\nu}_o)y_l^m(\so{})\thinspace\me^{i 2\pi \ro{}\cdot\bs{\nu}_o},\label{eq:objexp}
\end{align}
where $\mb{E}_l^m(\bs{\nu_o})$ and $\mb{M}_l^m(\bs{\nu_o})$ are the
vector spatio-angular spectra of the CSF and object, respectively, given by
\begin{align}
  \mb{E}_l^m(\bs{\nu_o}) &= \int_{\mathbb{R}^2}d\ro{} \int_{\mathbb{S}^2}d\so{}\, \tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{}) y_l^m(\so{})\thinspace\me^{-i 2\pi \ro{}\cdot\bs{\nu}_o},\\
\mb{M}_l^m(\bs{\nu_o}) &= \int_{\mathbb{R}^2}d\ro{} \int_{\mathbb{S}^2}d\so{}\, \bs{\mu}(\mb{r}_o, \so{}) y_l^m(\so{})\thinspace\me^{-i 2\pi \ro{}\cdot\bs{\nu}_o}.
\end{align}
In progress. By plugging Eqs. \ref{eq:csfexp} and \ref{eq:objexp} into
Eq. \ref{eq:eforward} we find that
\begin{align}
  \mb{e}_d(\rd{}) = \sum_{l=0}^{\infty} \sum_{m=-l}^l \int_{\mathbb{R}^2}d\bs{\nu}_o{}\, \mb{E}_l^m(\bs{\nu}_o)\mb{M}_l^m(\bs{\nu}_o)y_l^m(\so{})\thinspace\me^{i 2\pi (\rd{} - M\ro{})\cdot\bs{\nu}_o}. \label{eq:freqforward}
\end{align}
Eq. \ref{eq:freqforward} shows that the electric field on the detector can be
found by resolving the object into its spatio-angular components
$\mb{M}_l^m(\bs{\nu}_o)$, weighting each component by $\mb{E}_l^m(\bs{\nu}_o)$, then
summing over all spatio-angular components. For this reason, we identify
$\mb{E}_l^m(\bs{\nu}_o)$ as the spatio-angular \textit{coherent transfer function}
(CTF).

We define the spatio-angular optical transfer function (OTF) as
\begin{align}
  H_l^m(\bs{\nu}_o) = \int_{\mathbb{R}^3}d\ro{} \int_{\mathbb{S}^2}d\so{}\ h(\rd{}; \ro{}, \so{}) y_l^m(\so{})\thinspace\me^{i 2\pi \ro{}\cdot\bs{\nu}_o}. \label{eq:ft}
\end{align}
The spatio-angular OTF measures the ability of the microscope to pass
spatio-angular harmonics---instead of the usual spatial harmonics
$\me^{i 2\pi \ro{}\cdot\bs{\nu}_o}$ we now need consider the spatio-angular
harmonics $y_l^m(\so{})\thinspace\me^{i 2\pi
  \ro{}\cdot\bs{\nu}_o}$. Eq.~\ref{eq:ft} can be interpreted as the
spatio-angular Fourier transform of the spatio-angular PSF.
